---
title: "Getting Started with castSDM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with castSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**castSDM** (Causal-Aware Species Distribution Modeling) integrates
causal DAG learning with species distribution modeling to produce
ecologically interpretable and transferable predictions.

Unlike traditional SDMs that capture correlations, castSDM:

1. Discovers causal structure among environmental variables via DAGs
2. Uses the Markov Blanket for principled variable selection
3. Assigns causal roles (parent, child, co-parent) to selected variables
4. Trains multi-algorithm ensembles (RF, BRT, MaxEnt, GAM) for prediction
5. Projects habitat suitability under future climate scenarios

## Quick Start

```{r example, eval = FALSE}
library(castSDM)

# One-step pipeline
result <- cast(
  species_data = my_species_df,
  env_data = my_env_grid,
  models = c("rf", "brt", "maxent", "gam")
)

summary(result)
plot(result)
```

## Step-by-Step Usage

For more control, use individual pipeline functions:

```{r stepwise, eval = FALSE}
# 1. VIF screening
clean <- cast_vif(env_data, threshold = 10)

# 2. DAG learning (with presence as a node)
dag <- cast_dag(train_data, R = 100)

# 3. Variable selection (Markov Blanket + RF importance)
screen <- cast_select(dag, train_data)

# 4. Causal role assignment
roles <- cast_roles(screen, dag)

# 5. Model fitting
fit <- cast_fit(train_data, screen = screen, dag = dag,
                models = c("rf", "brt", "maxent", "gam"))

# 6. Spatial cross-validation
cv <- cast_cv(train_data, screen = screen, dag = dag,
              models = c("rf", "brt"))

# 7. Evaluation on holdout
eval <- cast_evaluate(fit, test_data)

# 8. Spatial prediction
pred <- cast_predict(fit, env_grid)

# 9. Ensemble prediction
ens <- cast_ensemble(fit, cv, env_grid)

# 10. Future projection (optional)
proj <- cast_project(fit, cv, current_env, future_envs)

# 11. CATE estimation (optional)
cate <- cast_cate(train_data, screen = screen, dag = dag, top_n = 3)
```
