---
title: "Getting Started with castSDM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with castSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**castSDM** (Causal-Aware Species Distribution Modeling) combines explicit
ecological role assumptions with fast covariate ranking and standard SDM
models. It does not prove ecological causality from observational data.

The active workflow:

1. Declares predictor roles before selection.
2. Allows only `direct_candidate` variables to enter the causal core.
3. Ranks eligible variables with shallow RF, invariance, conditional evidence,
   association, and redundancy control.
4. Reports the selected core and a complete role and exclusion audit.
5. Fits RF, BRT, MaxEnt, and GAM models and evaluates spatial transferability.
6. Optionally estimates CATE surfaces for selected variables.

## Quick Start

```{r example, eval = FALSE}
library(castSDM)

result <- cast(
  species_data = my_species_df,
  env_data = my_env_grid,
  models = c("rf", "brt", "maxent", "gam")
)

summary(result)
plot(result)
result$screen$roles
```

## Step-by-Step Usage

```{r stepwise, eval = FALSE}
# 1. Optional VIF screening
clean <- cast_vif(env_data, threshold = 10)

# 2. Review the default habitat role specification
vars <- get_env_vars(train_data)
spec <- cast_causal_spec(vars)
spec[spec$role != "direct_candidate", ]

# 3. Role-constrained variable screening
screen <- cast_select(data = train_data, causal_spec = spec)
screen$selected
screen$roles
screen$scores[, c("variable", "specified_role", "selection_tier",
                  "selection_reason")]

# 4. Model fitting
fit <- cast_fit(
  train_data,
  screen = screen,
  models = c("rf", "brt", "maxent", "gam")
)

# 5. Spatial cross-validation
cv <- cast_cv(
  train_data,
  screen = screen,
  models = c("rf", "brt")
)

# 6. Hold-out evaluation
eval <- cast_evaluate(fit, test_data)

# 7. Spatial prediction and ensemble
pred <- cast_predict(fit, env_grid)
ens <- cast_ensemble(fit, cv, env_grid, method = "weighted")

# 8. Future projection
proj <- cast_project(fit, cv, current_env, future_envs)

# 9. Optional CATE estimation
cate <- cast_cate(train_data, screen = screen, top_n = 3)
```

## Interpreting the Causal Audit

`cast_select()` returns a `cast_select` object. The selected variables are in
`screen$selected`, and the selected core is in `screen$roles`. The complete
audit, including excluded variables, is in `screen$scores`.

Only variables declared `direct_candidate` can be selected. Roles including
`mediator`, `proxy_only`, `sampling_bias`, `adjustment_only`, and `unknown`
remain visible but are ineligible. These declarations are reviewable ecological
assumptions, not confirmed mechanisms.
