---
title: "Beyond the defaults: tuning, weighting, replication, and diagnostics"
author: "castSDM authors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beyond the defaults: tuning, weighting, replication, and diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

The [getting-started vignette](castSDM.html) walks the default pipeline. This
vignette covers the options reviewers are most likely to ask about: per-engine
hyperparameter tuning, prevalence weighting, background designs beyond random
sampling, replicate pseudo-absence sets and replicated spatial
cross-validation, coordinate thinning, response curves, and change-map
coding that interchanges with biomod2.

Everything below slots into the same object model: `cast_prepare()` /
`cast_background()` produce data, `cast_fit()` produces a `cast_fit`,
`cast_cv()` produces a `cast_cv`, and the ensemble, projection and effect
products consume both. Nothing here changes the interpretation caveats in the
getting-started vignette; presence-background outputs remain relative
suitability on whatever reference the background defines.

```{r data}
library(castSDM)

data_file <- system.file("extdata",
  "CAST_Alces_alces_Res9_screened.csv", package = "castSDM")
dat <- read.csv(data_file)

set.seed(1)
dat <- dat[c(
  sample(which(dat$presence == 1), 150),
  sample(which(dat$presence == 0), 150)
), ]
```

## Coordinate thinning before anything else

Clumped records over-weight locally sampled areas and inflate apparent
performance under spatial CV. `cast_thin()` enforces a minimum great-circle
distance between retained records; the greedy pass visits records in a seeded
random order, so the seed decides which record of a conflicting cluster
survives and rows come back in their original order.

```{r thin}
dat <- cast_thin(dat, min_dist_km = 5, seed = 42)
```

Thinning is deliberately unforgiving about coordinates: projected metres
handed in as degrees are refused (a 10^5-fold thinning error, not a warning).

## Background designs: user-defined, bias-weighted, SRE, disk

`cast_background()` selects pseudo-absences by `strategy`. Beyond the default
random draw:

- `user_table` uses exactly the points you supply (Phillips et al. 2009
  user-defined background) — for already-available species records used as
  background, or expert-chosen controls.
- `bias_raster` draws cells with probability proportional to a raster of
  sampling effort (target-group background). Combine it only with
  `strategy = "random"`; bias weighting and the structured designs are
  competing designs, not layers.
- `strategy = "sre"` restricts the pool to the presence environmental
  envelope (each variable trimmed at `sre_quantile` on both sides;
  Barbet-Massin et al. 2012): geographically unconstrained, environmentally
  hard.
- `strategy = "disk"` restricts the pool to a distance band around
  presences: geographically close, hard pseudo-absences. Units follow the
  raster's map units, consistent with `buffer` in `cast_cv()`.

For `strategy = "environmental"` the stratification itself is now a choice:
`bin_method = "pca2d"` (default) bins a PC1 x PC2 grid; `bin_method =
"kmeans"` partitions the scaled environment space directly (the sdm
convention), which adapts better when drivers are correlated and the first
two PCs collapse them. `bin_k` sets the grid side or cluster count.

```{r background}
bg <- cast_background(dat, raster_stack = stk, strategy = "disk",
                      disk_min = 1, disk_max = 3, n_rep = 3, seed = 42)
```

`n_rep = 3` returns three independent pseudo-absence sets (a named
`cast_background_reps` list, `rep1`, `rep2`, ...). Fitting separately to
each replicate propagates PA-selection uncertainty instead of freezing one
arbitrary draw into the results. With `n_rep = 1` (default) the return is
the plain data.frame.

## Prevalence weighting

Presence-background fits describe a ratio you did not choose. When a target
prevalence matters for the application, `prevalence_target` weights the rows
so the fitted prevalence equals the target (weight = target/n per class, the
standard presence-background correction). RF, BRT and GAM accept the weights;
MaxEnt has no row-weight interface and stays unweighted. Evaluation metrics
remain unweighted — the target changes what is fitted, not what is scored.

```{r prevalence}
fit <- cast_fit(bg, models = c("rf", "brt", "gam", "maxent"),
                prevalence_target = 0.3, tune = TRUE, seed = 42)
```

## Hyperparameter tuning

`tune = TRUE` runs a small per-engine grid search: RF `mtry` scored by
out-of-bag TSS, BRT `interaction.depth` x `shrinkage` by internal-CV TSS,
MaxEnt feature classes x regularisation multiplier by stratified random-fold
TSS. The grids are informed by biomod2 and flexsdm and trimmed for cost;
combinations that fail to fit score `NA`, and a grid that fails entirely
keeps the fixed defaults with a warning.

Inside `cast_cv(tune = TRUE)` the grid runs again in every outer training
fold — nested, leak-free: the held-out folds never influence the chosen
hyperparameters. That multiplies compute by the grid size per fold; budget
accordingly.

## Replicated spatial cross-validation

`cast_cv()` blocks by space, so its metrics mix two uncertainty sources:
which rows landed in which fold, and where the block boundaries fell.
`n_repeat` separates them — each repeat redraws the fold assignment under
its own seed, metrics are averaged within each repeat, and the reported
standard deviation is the **between-repeat** spread (blocking randomness),
not fold noise.

```{r cv}
cv <- cast_cv(dat,
              models = c("rf", "brt"),
              select_method = "full",
              k = 4,
              n_repeat = 5,
              buffer = 50, buffer_unit = "km",
              tune = TRUE,
              prevalence_target = 0.3,
              threshold_method = "max_tss",
              seed = 42, verbose = FALSE)

cv$metrics[, c("auc_mean", "auc_sd", "tss_mean", "tss_sd")]
```

The buffer exclusion band also takes units now: `buffer_unit = "km"`
measures great-circle kilometres (Haversine) and requires decimal-degree
coordinates; `"deg"` keeps the planar-degree behaviour of previous
versions. `fold_metrics` gains a `replicate` column, `selection_freq` pools
all repeats, and the out-of-fold predictions refer to the first repeat.

## Response curves

`cast_response_curves()` traces how predicted suitability moves along one
driver (`type = "fixed"`: other predictors at training medians) or averaged
over them (`type = "pdp"`: partial dependence, Friedman 2001, with the
background rows subsampled deterministically at 2000 for cost). A `pair`
argument adds two-variable interaction grids. The `plot()` method handles
both layouts.

```{r curves}
rc <- cast_response_curves(fit, cv = cv,
                           variables = c("bio01", "bio12"),
                           type = "pdp")
plot(rc)

rc_pair <- cast_response_curves(fit, cv = cv,
                                pair = c("bio01", "bio12"))
plot(rc_pair)
```

Read curves on the drivers the selection admitted; a curve along a dropped
variable describes the model, not the species.

## Projection with interchangeable change coding

`cast_project()` and `cast_project_raster()` write change maps in either the
package's internal encoding (`gain=1, loss=-1, stable_present=2,
stable_absent=0`) or biomod2's `bm_RangeSize()` convention (`Gain=1,
Loss=-2, Stable_Pres=-1, Stable_Abs=0`) via `coding = "biomod2"`, so saved
GeoTIFFs feed BIOMOD_RangeSize visualisation scripts without remapping.
Projection statistics report `n_current_range` / `n_future_range` pixel
counts and the biomod2-convention `pct_loss` (loss over the current range)
and `pct_gain` (gain over the future range).

```{r projection}
proj <- cast_project(fit, cv, current_env, future_envs,
                     coding = "biomod2")
proj$stats
```

## What the extra options do not buy

Tuning, weighting, replication and buffers sharpen the model and its
uncertainty bands; none of them identifies an adjustment set, detects
unobserved confounding, or converts relative suitability to occurrence
probability. The masking and interpretation notes in the getting-started
vignette apply unchanged to every product shown here.
