---
title: "Getting Started with castSDM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with castSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**castSDM** (Causal Structure-informed Species Distribution Modeling) integrates
causal inference methods with species distribution modeling to produce
ecologically interpretable and transferable predictions.

Unlike traditional SDMs that capture correlations, castSDM:

1. Discovers causal structure among environmental variables via DAGs
2. Estimates causal effect sizes using Double Machine Learning
3. Engineers features guided by causal theory
4. Trains causally-informed neural networks for prediction
5. Maps spatially heterogeneous treatment effects

## Quick Start

```{r example, eval = FALSE}
library(castSDM)

# One-step pipeline
result <- cast(
  species_data = my_species_df,
  env_data = my_env_grid,
  models = c("cast", "rf", "maxent", "brt")
)

summary(result)
plot(result)
```

## Step-by-Step Usage

For more control, use individual pipeline functions:

```{r stepwise, eval = FALSE}
# 1. VIF screening
clean <- cast_vif(env_data, threshold = 10)

# 2. DAG learning
dag <- cast_dag(train_data, R = 100)

# 3. ATE estimation
ate <- cast_ate(train_data, K = 2)

# 4. Variable screening
vars <- cast_screen(dag, ate, train_data)

# 5. Causal role assignment
roles <- cast_roles(vars, dag)

# 6. Model fitting (feature engineering is handled internally)
fit <- cast_fit(train_data, screen = vars, dag = dag, ate = ate,
                models = c("cast", "rf"))

# 7. Evaluation
eval <- cast_evaluate(fit, test_data)

# 8. Prediction
pred <- cast_predict(fit, env_grid)

# 9. CATE estimation
cate <- cast_cate(train_data, ate = ate, top_n = 3)
```
