---
title: "Getting started with castSDM"
author: "castSDM authors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with castSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

`castSDM` is an end-to-end species distribution modelling (SDM) toolkit whose
variable-selection stage is built on conditional screening rather than
associational filters:

- `cast_select(method = "cpi")` (default) keeps predictors whose conditional
  predictive impact survives Benjamini-Hochberg FDR control, testing each
  predictor *given every other predictor*;
- `cast_select(method = "dml")` reports signed Neyman-orthogonal
  partial-linear effects via `cast_effect()`;
- selection is re-run inside every spatial cross-validation fold, so
  held-out data never influence variable choice.

The rest of the pipeline is standard SDM engineering: RF / BRT / MaxEnt / GAM
learners, nested spatial CV, MESS extrapolation flags, performance-weighted
ensembles (N-SDM score), ESM for rare species, and CMIP6 projection.

## Data

Species data are presence/background tables with `lon`, `lat`, `presence`
(0/1) and environmental predictors. `inst/extdata` ships a small example
table (a 2000-row subset of the full screening table):

```{r data}
library(castSDM)

data_file <- system.file("extdata",
  "CAST_Alces_alces_Res9_screened.csv", package = "castSDM")
dat <- read.csv(data_file)

# Keep the workflow light for this vignette: subsample rows.
set.seed(1)
dat <- dat[c(
  sample(which(dat$presence == 1), 150),
  sample(which(dat$presence == 0), 150)
), ]

head(dat[, c("lon", "lat", "presence", "bio01", "bio12", "elevation")])
```

## One-step pipeline

```{r pipeline}
result <- cast(
  dat,
  models = c("rf", "gam"),
  select_method = "cpi",
  select_min_vars = 3,
  select_num_trees = 100,
  do_cv = TRUE,
  cv_k = 3,
  seed = 42,
  verbose = FALSE
)

summary(result)
plot(result$screen)
```

## Signed effect reporting with DML

```{r effect}
screen_dml <- cast_select(dat, method = "dml", num_trees = 100,
                          dml_folds = 3, seed = 42, verbose = FALSE)
eff <- cast_effect(screen_dml)
plot(eff)
```

## Counterfactual what-if map

```{r counterfactual}
# Shift the first predictor retained by the screen (guaranteed to be fitted).
cf <- cast_counterfactual(result$fit, newdata = dat,
                          variable = result$fit$env_vars[1], shift = 1,
                          shift_type = "sd")
plot(cf, basemap = "none")
```

## Spatial prediction on a grid

```{r predict}
# Build a small prediction grid from the data range.
grid <- expand.grid(
  lon = seq(min(dat$lon), max(dat$lon), length.out = 20),
  lat = seq(min(dat$lat), max(dat$lat), length.out = 20)
)
env_vars <- result$fit$env_vars
for (v in env_vars) {
  grid[[v]] <- median(dat[[v]], na.rm = TRUE)
}

pred <- cast_predict(result$fit, grid)
plot(pred, basemap = "none")
```

## Rare species: ensemble of small models

```{r esm}
rare <- dat[1:60, ]
rare$presence[1:50] <- 0   # 10 presences, 50 backgrounds
esm <- cast_esm(rare, top_k = 4, base_algo = "glm", seed = 7, verbose = FALSE)
print(esm)
```

## Batch + config

```{r batch}
template <- cast_config_template(tempfile(fileext = ".yaml"), overwrite = TRUE)
batch <- cast_batch(
  list("example" = dat),
  models = "rf",
  do_cv = TRUE, cv_k = 3,
  select_method = "rf", select_num_trees = 100,
  parallel = FALSE, seed = 1, verbose = FALSE
)
print(batch)
```
