---
title: "Getting Started with castSDM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with castSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**castSDM** (Causal-Aware Species Distribution Modeling) combines
response-focused Markov Blanket screening with standard SDM models. Its main
use is not to prove ecological causality from observational data, but to make
variable screening more explicit, sparse, and reproducible.

The active workflow:

1. Learns a response-focused DAG or MB screening graph.
2. Forbids `presence -> environmental predictor` edges by default.
3. Selects variables with Markov Blanket membership plus RF importance.
4. Reports screening roles:
   `mb_direct`, `mb_associated`, `importance_added`, and
   `importance_screened`.
5. Fits RF, BRT, MaxEnt, and GAM models and evaluates spatial transferability.
6. Optionally estimates CATE surfaces for selected variables.

## Quick Start

```{r example, eval = FALSE}
library(castSDM)

result <- cast(
  species_data = my_species_df,
  env_data = my_env_grid,
  models = c("rf", "brt", "maxent", "gam"),
  dag_structure_method = "mb_first",
  dag_response_as_sink = TRUE
)

summary(result)
plot(result)
result$screen$roles
```

## Step-by-Step Usage

```{r stepwise, eval = FALSE}
# 1. Optional VIF screening
clean <- cast_vif(env_data, threshold = 10)

# 2. DAG / MB screening graph
dag <- cast_dag(
  train_data,
  structure_method = "mb_first",
  include_response = TRUE,
  response_as_sink = TRUE
)

# 3. Variable screening
screen <- cast_select(dag, train_data)
screen$selected
screen$roles

# 4. Model fitting
fit <- cast_fit(
  train_data,
  screen = screen,
  dag = dag,
  models = c("rf", "brt", "maxent", "gam")
)

# 5. Spatial cross-validation
cv <- cast_cv(
  train_data,
  screen = screen,
  dag = dag,
  models = c("rf", "brt")
)

# 6. Hold-out evaluation
eval <- cast_evaluate(fit, test_data)

# 7. Spatial prediction and ensemble
pred <- cast_predict(fit, env_grid)
ens <- cast_ensemble(fit, cv, env_grid, method = "weighted")

# 8. Future projection
proj <- cast_project(fit, cv, current_env, future_envs)

# 9. Optional CATE estimation
cate <- cast_cate(train_data, screen = screen, dag = dag, top_n = 3)
```

## Interpreting Screening Roles

`cast_select()` returns a `cast_select` object. The selected variables are in
`screen$selected`, and their screening roles are in `screen$roles`.

The roles mean:

- `mb_direct`: adjacent to the response in the response-focused MB graph.
- `mb_associated`: retained through broader MB structure.
- `importance_added`: outside the MB but retained by RF importance.
- `importance_screened`: retained by RF importance when MB membership is dense
  or otherwise non-selective.

These roles describe the screening path for each variable. They should not be
read as confirmed ecological mechanisms.
