---
title: "Getting Started with castSDM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with castSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**castSDM** (Causal-Aware Species Distribution Modeling) combines
spatially invariant causal screening with standard SDM models. Its main
use is not to prove ecological causality from observational data, but to make
variable screening more explicit, sparse, transferable, and reproducible.

The active workflow:

1. Learns a response-focused DAG or MB screening graph.
2. Forbids `presence -> environmental predictor` edges by default.
3. Selects variables with RF importance, bootstrap stability, spatial-block
   effect consistency, and correlation-cluster redundancy control.
4. Reports screening roles:
   `invariant_driver`, `stable_predictive`, `predictive_rescue`, and
   `redundant_proxy`.
5. Fits RF, BRT, MaxEnt, and GAM models and evaluates spatial transferability.
6. Optionally estimates CATE surfaces for selected variables.

## Quick Start

```{r example, eval = FALSE}
library(castSDM)

result <- cast(
  species_data = my_species_df,
  env_data = my_env_grid,
  models = c("rf", "brt", "maxent", "gam"),
  dag_structure_method = "mb_first",
  dag_response_as_sink = TRUE
)

summary(result)
plot(result)
result$screen$roles
```

## Step-by-Step Usage

```{r stepwise, eval = FALSE}
# 1. Optional VIF screening
clean <- cast_vif(env_data, threshold = 10)

# 2. DAG / MB screening graph
dag <- cast_dag(
  train_data,
  structure_method = "mb_first",
  include_response = TRUE,
  response_as_sink = TRUE
)

# 3. Variable screening
screen <- cast_select(dag, train_data)
screen$selected
screen$roles

# 4. Model fitting
fit <- cast_fit(
  train_data,
  screen = screen,
  dag = dag,
  models = c("rf", "brt", "maxent", "gam")
)

# 5. Spatial cross-validation
cv <- cast_cv(
  train_data,
  screen = screen,
  dag = dag,
  models = c("rf", "brt")
)

# 6. Hold-out evaluation
eval <- cast_evaluate(fit, test_data)

# 7. Spatial prediction and ensemble
pred <- cast_predict(fit, env_grid)
ens <- cast_ensemble(fit, cv, env_grid, method = "weighted")

# 8. Future projection
proj <- cast_project(fit, cv, current_env, future_envs)

# 9. Optional CATE estimation
cate <- cast_cate(train_data, screen = screen, dag = dag, top_n = 3)
```

## Interpreting Screening Roles

`cast_select()` returns a `cast_select` object. The selected variables are in
`screen$selected`, and their screening roles are in `screen$roles`.

The roles mean:

- `invariant_driver`: selected because it is predictive, direction-consistent,
  and spatially stable.
- `stable_predictive`: selected because it is predictive and stable, with
  weaker effect-direction evidence.
- `predictive_rescue`: retained to preserve model performance.
- `redundant_proxy`: rejected because a higher-scoring correlated variable was
  already selected.

These roles describe the screening path for each variable. They should not be
read as confirmed ecological mechanisms.
