---
title: "Getting started with castSDM"
author: "castSDM authors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with castSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

`castSDM` is an end-to-end species distribution modelling (SDM) toolkit with
two separate concerns kept separate:

- **Variable selection for parsimony**, not for attribution.
  `cast_select()` runs a two-stage screen: collinearity thinning at
  \|r\| <= 0.7 (Dormann et al. 2013), then an importance filter against a
  permutation null (Altmann et al. 2010). Selection is re-run inside every
  spatial cross-validation fold, so held-out data never influence variable
  choice.
- **Attribution as an audited pair.** `cast_effect_table()` and
  `cast_effect_map()` report the interventional response to each driver
  (shift the driver, hold every other driver fixed), and `cast_necessity()`
  reports what dropping the driver costs in held-out AUC. A driver with a
  large effect but no knockout cost is substitutable by a collinear partner,
  so attribution to it is not identified. Report both.

The rest of the pipeline is standard SDM engineering: RF / BRT / MaxEnt / GAM
learners, nested spatial CV, MESS extrapolation flags, performance-weighted
ensembles (N-SDM score), and CMIP6 projection.

## Data

Species data are presence/background tables with `lon`, `lat`, `presence`
(0/1) and environmental predictors. `inst/extdata` ships a small example
table (a 2000-row subset of the full screening table):

```{r data}
library(castSDM)

data_file <- system.file("extdata",
  "CAST_Alces_alces_Res9_screened.csv", package = "castSDM")
dat <- read.csv(data_file)

# Keep the workflow light for this vignette: subsample rows.
set.seed(1)
dat <- dat[c(
  sample(which(dat$presence == 1), 150),
  sample(which(dat$presence == 0), 150)
), ]

head(dat[, c("lon", "lat", "presence", "bio01", "bio12", "elevation")])
```

## One-step pipeline

```{r pipeline}
result <- cast(
  dat,
  models = c("rf", "gam"),
  select_method = "two_stage",
  select_num_trees = 100,
  select_n_perm = 19,       # small example: a short permutation null
  do_cv = TRUE,
  cv_k = 3,
  seed = 42,
  verbose = FALSE
)

summary(result)
plot(result$screen)
```

## Attribution: effect size and necessity

`cast_effect_table()` reports the interventional response to each driver, and
`cast_necessity()` reports what dropping it costs out of sample. Read them
together.

```{r attribution}
eff <- cast_effect_table(result$fit, newdata = dat, verbose = FALSE)
eff        # rank by mean_abs_dHSS; mean_signed_dHSS gives the direction

nec <- cast_necessity(dat, screen = result$screen, k = 3, num_trees = 100,
                      seed = 42, verbose = FALSE)
nec        # mean_dAUC: held-out AUC lost by dropping each driver
```

A driver with a large `mean_abs_dHSS` but `mean_dAUC` near zero is
substitutable by a collinear partner: the model responds to it, but the data
cannot attribute habitat change to *that* driver specifically.

## Sensitivity what-if map

```{r sensitivity}
# Shift the first predictor retained by the screen (guaranteed to be fitted).
cf <- cast_sensitivity(result$fit, newdata = dat,
                       variable = result$fit$env_vars[1], shift = 1,
                       shift_type = "sd")
plot(cf, basemap = "none")
```

## Spatial prediction on a grid

```{r predict}
# Build a small prediction grid from the data range.
grid <- expand.grid(
  lon = seq(min(dat$lon), max(dat$lon), length.out = 20),
  lat = seq(min(dat$lat), max(dat$lat), length.out = 20)
)
env_vars <- result$fit$env_vars
for (v in env_vars) {
  grid[[v]] <- median(dat[[v]], na.rm = TRUE)
}

pred <- cast_predict(result$fit, grid)
plot(pred, basemap = "none")
```
