---
title: "Getting started with castSDM"
author: "castSDM authors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with castSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

`castSDM` is an end-to-end species distribution modelling (SDM) toolkit built
around one question: if a driver moved, what would happen to predicted
suitability, and where is that answer supported by data?

Two concerns stay separate:

- **Default selection ranks predictors by conditional changes in the fitted
  response.** `cast_select()` runs a two-stage screen: collinearity thinning
  at \|r\| <= 0.7 (Dormann et al. 2013), then a filter against a
  within-stratum permutation null using the **conditional effect** — how far
  the fitted probability moves when the predictor is shifted and every other
  predictor stays at its observed value. Shuffling only among rows with
  similar values on the other predictors keeps the null inside the observed
  joint support (Hooker, Mentch & Zhou 2021). Selection is re-run inside
  every spatial cross-validation fold, so held-out data never influence
  variable choice.
- **One shift, one map, hard masking.** The shift response to each driver is
  reported as a table and a spatial map for a single raw-unit shift (for
  example `bio01 + 2` degrees). Rows or cells outside the training quantile
  box — at baseline or after shifting — are masked, not ranked.

The rest of the pipeline is standard SDM engineering: RF / BRT / MaxEnt / GAM
learners, nested spatial CV, MESS extrapolation flags, performance-weighted
ensembles (N-SDM score), and CMIP6 projection.

## Data

Species data are presence/background tables with `lon`, `lat`, `presence`
(0/1) and environmental predictors. `inst/extdata` ships a small example
table (a 2000-row subset of the full screening table):

```{r data}
library(castSDM)

data_file <- system.file("extdata",
  "CAST_Alces_alces_Res9_screened.csv", package = "castSDM")
dat <- read.csv(data_file)

# Keep the workflow light for this vignette: subsample rows.
set.seed(1)
dat <- dat[c(
  sample(which(dat$presence == 1), 150),
  sample(which(dat$presence == 0), 150)
), ]

head(dat[, c("lon", "lat", "presence", "bio01", "bio12", "elevation")])
```

## One-step pipeline

```{r pipeline}
result <- cast(
  dat,
  models = c("rf", "gam"),
  select_method = "two_stage",
  select_num_trees = 100,
  select_n_perm = 19,       # small example: a short conditional null
  do_cv = TRUE,
  cv_k = 3,
  seed = 42,
  verbose = FALSE
)

summary(result)
plot(result$screen)
```

## What each stage measured

`cast_importance()` shows the single conditional-effect column from the
screen. Predictors above their own within-stratum null are retained, up to
the sample-size cap.

```{r importance}
imp <- cast_importance(result$screen)
imp
plot(imp)
```

## Shift effect and masking

```{r attribution}
eff <- cast_effect_table(result$fit, newdata = dat, shift = 1,
                         verbose = FALSE)
eff        # rank supported drivers by mean_abs_dHSS; masked drivers read NA
plot(eff)  # cross = masked: the shift is not answerable inside the box

# The driver the screen ranked first is guaranteed to be fitted.
v1 <- result$screen$scores$variable[which.max(
  result$screen$scores$interventional_effect)]
```

Give the shift in raw units whenever it has ecological meaning, for example
`shift = list(bio01 = 2)` for +2 degrees. A shift that leaves the training
range everywhere is refused rather than answered.

## Spatial prediction on a grid

```{r predict}
# Build a small prediction grid from the data range.
grid <- expand.grid(
  lon = seq(min(dat$lon), max(dat$lon), length.out = 20),
  lat = seq(min(dat$lat), max(dat$lat), length.out = 20)
)
env_vars <- result$fit$env_vars
for (v in env_vars) {
  grid[[v]] <- median(dat[[v]], na.rm = TRUE)
}

pred <- cast_predict(result$fit, grid)
plot(pred, basemap = "none")
```

## What the numbers are not

Every effect product is a **model-based shift contrast**
(g-computation / standardization). A causal interpretation requires consistency
of the stated shift, a justified adjustment set, no uncontrolled confounding,
joint positivity, and adequate response and observation models. Masking checks
only training-quantile-box coverage of baseline and shifted rows: it cannot
detect unobserved combinations inside the box or establish positivity.
Automatic predictor selection does not identify an adjustment set. The
estimator is not doubly robust. On presence/background data the scale
is relative suitability, not occurrence probability.
