---
title: "Getting started with castSDM"
author: "castSDM authors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with castSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE)
```

## Overview

`castSDM` is an end-to-end species distribution modelling (SDM) toolkit whose
variable-selection stage is built on conditional screening rather than
associational filters:

- `cast_select(method = "cpi")` (default) keeps predictors whose conditional
  predictive impact survives Benjamini-Hochberg FDR control, testing each
  predictor *given every other predictor*;
- `cast_importance()` turns that screen into a tidy per-predictor table with
  confidence intervals, and `cast_sensitivity()` maps single-driver what-if
  shifts;
- selection is re-run inside every spatial cross-validation fold, so
  held-out data never influence variable choice.

The rest of the pipeline is standard SDM engineering: RF / BRT / MaxEnt / GAM
learners, nested spatial CV, MESS extrapolation flags, performance-weighted
ensembles (N-SDM score), and CMIP6 projection.

## Data

Species data are presence/background tables with `lon`, `lat`, `presence`
(0/1) and environmental predictors. `inst/extdata` ships a small example
table (a 2000-row subset of the full screening table):

```{r data}
library(castSDM)

data_file <- system.file("extdata",
  "CAST_Alces_alces_Res9_screened.csv", package = "castSDM")
dat <- read.csv(data_file)

# Keep the workflow light for this vignette: subsample rows.
set.seed(1)
dat <- dat[c(
  sample(which(dat$presence == 1), 150),
  sample(which(dat$presence == 0), 150)
), ]

head(dat[, c("lon", "lat", "presence", "bio01", "bio12", "elevation")])
```

## One-step pipeline

```{r pipeline}
result <- cast(
  dat,
  models = c("rf", "gam"),
  select_method = "cpi",
  select_num_trees = 100,
  select_min_vars = 3,      # small example: floor the screen at 3 predictors
  do_cv = TRUE,
  cv_k = 3,
  seed = 42,
  verbose = FALSE
)

summary(result)
plot(result$screen)
```

## Conditional importance

```{r importance}
eff <- cast_importance(result)               # tidy CPI table + CIs
plot(eff)                                    # coefficient (forest) plot
```

## Sensitivity what-if map

```{r sensitivity}
# Shift the first predictor retained by the screen (guaranteed to be fitted).
cf <- cast_sensitivity(result$fit, newdata = dat,
                       variable = result$fit$env_vars[1], shift = 1,
                       shift_type = "sd")
plot(cf, basemap = "none")
```

## Spatial prediction on a grid

```{r predict}
# Build a small prediction grid from the data range.
grid <- expand.grid(
  lon = seq(min(dat$lon), max(dat$lon), length.out = 20),
  lat = seq(min(dat$lat), max(dat$lat), length.out = 20)
)
env_vars <- result$fit$env_vars
for (v in env_vars) {
  grid[[v]] <- median(dat[[v]], na.rm = TRUE)
}

pred <- cast_predict(result$fit, grid)
plot(pred, basemap = "none")
```
