---
title: "Getting started with castSDM"
author: "castSDM authors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with castSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

`castSDM` is an end-to-end species distribution modelling (SDM) toolkit built
around one question: if a driver moved, what would happen to predicted
suitability, and where is that answer supported by data?

Two concerns stay separate:

- **Default selection is a spatial forward search.** `cast_select()` runs a
  two-stage screen: collinearity thinning at \|r\| <= 0.7 (Dormann et al.
  2013), then forward selection on the inner cross-validated loss of a
  probability random forest — spatial folds when coordinates exist. Each
  addition must improve held-out loss beyond a paired 2-SE guard, so the
  search stops by itself: no predictor-count cap, no fallback set, and an
  empty selection honestly means nothing beat the intercept model (Meyer et
  al. 2018, 2019). Selection is re-run inside every spatial
  cross-validation fold, so held-out data never influence variable choice.
- **One shift, one map, hard masking.** The shift response to each driver is
  reported as a table and a spatial map for a single raw-unit shift (for
  example `bio01 + 2` degrees). Rows or cells outside the training quantile
  box — at baseline or after shifting — are masked, not ranked.

The rest of the pipeline is standard SDM engineering: RF / BRT / MaxEnt / GAM
learners, nested spatial CV, MESS extrapolation flags, performance-weighted
ensembles (N-SDM score), and CMIP6 projection.

## Data

Species data are presence/background tables with `lon`, `lat`, `presence`
(0/1) and environmental predictors. `inst/extdata` ships a small example
table (a 2000-row subset of the full screening table):

```{r data}
library(castSDM)

data_file <- system.file("extdata",
  "CAST_Alces_alces_Res9_screened.csv", package = "castSDM")
dat <- read.csv(data_file)

# Keep the workflow light for this vignette: subsample rows.
set.seed(1)
dat <- dat[c(
  sample(which(dat$presence == 1), 150),
  sample(which(dat$presence == 0), 150)
), ]

head(dat[, c("lon", "lat", "presence", "bio01", "bio12", "elevation")])
```

## One-step pipeline

```{r pipeline}
result <- cast(
  dat,
  models = c("rf", "gam"),
  select_method = "two_stage",
  select_num_trees = 100,
  select_n_folds = 3,     # small example: a short inner search
  do_cv = TRUE,
  cv_k = 3,
  seed = 42,
  verbose = FALSE
)

summary(result)
plot(result$screen)
```

## What each stage measured

`cast_importance()` shows the forward-selection path from the screen:
each admitted predictor with its admission step and inner-CV loss
improvement. There are no p-values and no count cap to read around.

```{r importance}
imp <- cast_importance(result$screen)
imp
plot(imp)
```

## Shift effect and masking

```{r attribution}
eff <- cast_effect_table(result$fit, newdata = dat, shift = 1,
                         verbose = FALSE)
eff        # rank supported drivers by mean_abs_dHSS; masked drivers read NA
plot(eff)  # cross = masked: the shift is not answerable inside the box

# The first-admitted driver is guaranteed to be fitted.
v1 <- result$screen$selected[1]
```

Give the shift in raw units whenever it has ecological meaning, for example
`shift = list(bio01 = 2)` for +2 degrees. A shift that leaves the training
range everywhere is refused rather than answered.

## Spatial prediction on a grid

```{r predict}
# Build a small prediction grid from the data range.
grid <- expand.grid(
  lon = seq(min(dat$lon), max(dat$lon), length.out = 20),
  lat = seq(min(dat$lat), max(dat$lat), length.out = 20)
)
env_vars <- result$fit$env_vars
for (v in env_vars) {
  grid[[v]] <- median(dat[[v]], na.rm = TRUE)
}

pred <- cast_predict(result$fit, grid)
plot(pred, basemap = "none")
```

## What the numbers are not

Every effect product is a **model-based shift contrast**
(g-computation / standardization). A causal interpretation requires consistency
of the stated shift, a justified adjustment set, no uncontrolled confounding,
joint positivity, and adequate response and observation models. Masking checks
only training-quantile-box coverage of baseline and shifted rows: it cannot
detect unobserved combinations inside the box or establish positivity.
Automatic predictor selection does not identify an adjustment set. The
estimator is not doubly robust. On presence/background data the scale
is relative suitability, not occurrence probability.
