---
title: "Getting started with castSDM"
author: "castSDM authors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with castSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

`castSDM` is an end-to-end species distribution modelling (SDM) toolkit built
around one question: if a driver moved, what would happen to predicted
suitability, and where is that answer supported by data?

Two concerns stay separate:

- **Default selection ranks predictors by changes in the fitted response.**
  `cast_select()` runs a two-stage screen: collinearity thinning at
  \|r\| <= 0.7 (Dormann et al. 2013), then a filter against a
  permuted-response null using the **interventional effect** — how far the
  fitted probability moves when the predictor is shifted and every other
  predictor stays at its observed value. The same forest's permutation
  importance is reported alongside, because permuting a predictor breaks its
  correlation with the others and is therefore governed by the model's
  extrapolation behaviour (Hooker, Mentch & Zhou 2021). Selection is re-run
  inside every spatial cross-validation fold, so held-out data never influence
  variable choice.
- **Effects come with their response shape and their support.** The
  interventional response to each driver is reported as a table, a spatial map
  and a dose-response curve, together with the fraction of observed predictor
  vectors still inside the training support after the shift.
  `cast_necessity()` separately reports what dropping a driver costs in
  held-out AUC.

For prespecified, scientifically justified environment groups,
`cast_select(method = "tramicp", environment = "environment_group")` instead
uses the optional tramicp package to test all candidate subsets for invariance.
It does not apply the two-stage screen first, fabricate groups from CV folds,
or replace an empty intersection with a ranked set. Valid environments and
model/sampling assumptions remain necessary; the result is not a verified
causal adjustment set.

For a single fitted GAM, `cast_sensitivity(backend = "marginaleffects",
model = "gam", ...)` provides explicit baseline-versus-shifted comparisons.
The native backend remains available for multiple models. Neither backend turns
presence/background suitability into occurrence probability or identifies a
causal effect merely by holding the other predictors fixed.

The rest of the pipeline is standard SDM engineering: RF / BRT / MaxEnt / GAM
learners, nested spatial CV, MESS extrapolation flags, performance-weighted
ensembles (N-SDM score), and CMIP6 projection.

## Data

Species data are presence/background tables with `lon`, `lat`, `presence`
(0/1) and environmental predictors. `inst/extdata` ships a small example
table (a 2000-row subset of the full screening table):

```{r data}
library(castSDM)

data_file <- system.file("extdata",
  "CAST_Alces_alces_Res9_screened.csv", package = "castSDM")
dat <- read.csv(data_file)

# Keep the workflow light for this vignette: subsample rows.
set.seed(1)
dat <- dat[c(
  sample(which(dat$presence == 1), 150),
  sample(which(dat$presence == 0), 150)
), ]

head(dat[, c("lon", "lat", "presence", "bio01", "bio12", "elevation")])
```

## One-step pipeline

```{r pipeline}
result <- cast(
  dat,
  models = c("rf", "gam"),
  select_method = "two_stage",
  select_num_trees = 100,
  select_n_perm = 19,       # small example: a short permutation null
  do_cv = TRUE,
  cv_k = 3,
  seed = 42,
  verbose = FALSE
)

summary(result)
plot(result$screen)
```

## What each stage measured

`cast_importance()` shows both attribution columns from the screen and how far
they agree. A predictor the forest relies on but barely responds to when
shifted is a collinear stand-in; the rank agreement summarises how much the
choice of statistic mattered here.

```{r importance}
imp <- cast_importance(result$screen)
imp
plot(imp)
```

## Effect size, response shape and support

```{r attribution}
eff <- cast_effect_table(result$fit, newdata = dat, verbose = FALSE)
eff        # rank by mean_abs_dHSS; mean_signed_dHSS gives the direction
plot(eff)  # hollow = shift answered mostly by extrapolation (support < 0.5)

# The driver the screen ranked first is guaranteed to be fitted.
v1 <- result$screen$scores$variable[which.max(
  result$screen$scores$interventional_effect)]

# Shape: how the effect scales with the size of the shift.
dr <- cast_dose_response(result$fit, v1, shift = seq(-2, 2, by = 0.5))
plot(dr)

# Support: where the shift is answerable at all.
sp <- cast_effect_support(result$fit, variables = v1, shift = c(1, 2, 3))
plot(sp)
```

A dose-response point drawn hollow is answered mostly by extrapolation.
`cast_effect_support()` reports the same information as a number, and a shift
that leaves the training range everywhere is refused rather than answered.

## Necessity audit

```{r necessity}
nec <- cast_necessity(dat, screen = result$screen, k = 3, num_trees = 100,
                      seed = 42, verbose = FALSE)
nec        # mean_dAUC: held-out AUC lost by dropping each driver
plot(nec)  # whiskers = fold range; read beside plot(eff)
```

A driver with a large `mean_abs_dHSS` but `mean_dAUC` near zero shows a model
response with limited RF knockout cost. Redundancy, limited power, AUC and
estimator choice can all contribute. Neither diagnostic is a causal test on
its own.

## Sensitivity what-if map

```{r sensitivity}
cf <- cast_sensitivity(result$fit, newdata = dat, variable = v1, shift = 1,
                       shift_type = "sd")
plot(cf, basemap = "none")
```

## Spatial prediction on a grid

```{r predict}
# Build a small prediction grid from the data range.
grid <- expand.grid(
  lon = seq(min(dat$lon), max(dat$lon), length.out = 20),
  lat = seq(min(dat$lat), max(dat$lat), length.out = 20)
)
env_vars <- result$fit$env_vars
for (v in env_vars) {
  grid[[v]] <- median(dat[[v]], na.rm = TRUE)
}

pred <- cast_predict(result$fit, grid)
plot(pred, basemap = "none")
```

## What the numbers are not

Every effect product is a **model-based interventional contrast**
(g-computation / standardization). A causal interpretation requires consistency,
a justified adjustment set, no uncontrolled confounding, joint positivity,
and adequate response and observation models. The support column checks only
training-quantile-box coverage of baseline and shifted rows: it cannot detect
unobserved combinations inside the box or establish positivity. The
`range_supported` flag is a descriptive coverage threshold, not causal
estimability. Automatic predictor selection does not identify an adjustment
set. The estimator is not doubly robust. On presence/background data the scale
is relative suitability, not occurrence probability.
