---
title: "YAML-driven Workflow for Multi-species Runs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{YAML-driven Workflow for Multi-species Runs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Why a YAML config?

Reproducibility for multi-species batches lives or dies on parameter
discipline. A single `cast_batch()` call already carries dozens of named
arguments; tracking which seed, fold count, or DAG threshold was used for
which run quickly becomes a chore.

`cast_run_from_config()` makes the parameter set the *artifact*. Save it
alongside your data, version it with git, and a year later you can re-run
the analysis with one line. The same YAML doubles as the documentation of
the run.

## Quick start

Generate a starter template and edit it:

```{r}
library(castSDM)
cast_config_template("my_run.yaml")
```

The generated file (also bundled at
`system.file("templates", "castSDM_config.yaml.tmpl", package = "castSDM")`)
is annotated; the bare-minimum schema is:

```yaml
run:
  output_dir: "castSDM_output"
  seed: 1234
  parallel: true
  worker_budget:
    total_workers: 8
  resume: true

species:
  species_a:
    file: "data/species_a.csv"
  species_b:
    file: "data/species_b.csv"

models: [cast, rf, maxent, brt]
backend: r            # pure-R CI-MLP (no torch needed)

cast_batch_args:
  train_fraction: 0.7
  do_cv: true
  cv_k: 5
```

Each `species.<name>.file` CSV must contain at least `lon`, `lat`,
`presence`, plus any environmental columns. Paths are resolved relative
to the directory containing the YAML file.

Run the pipeline:

```{r}
out <- cast_run_from_config("my_run.yaml")
```

`out` is a `cast_batch` object — same as if you had called
`cast_batch()` directly.

## Resuming after a crash

Long batches can be interrupted by anything from a kernel panic to a
power loss. The toolkit writes per-step checkpoints under
`output_dir/<species>/.steps/` (DAG, ATE, screen, fit, CV, predict, CATE,
consistency) and a per-species `cast_result.rds` once everything is done.

When `run.resume: true` is set in the YAML and any species in
`output_dir` has a finished `cast_result.rds`, the helper transparently
delegates to `cast_batch_resume()`. Already-complete species are skipped;
species that were *partially* finished load their cached steps and pick
up where they left off.

To recover manually:

```{r}
cast_batch_resume(
  output_dir   = "castSDM_output",
  species_list = list(species_a = df_a, species_b = df_b),
  models       = c("cast", "rf", "maxent", "brt")
)
```

## Worker budget on Windows

Set `run.worker_budget.total_workers` to keep the CPU busy without
oversubscribing cores. Internally this calls `cast_worker_budget()`,
which splits workers across two nested layers:

* the species fan-out, and
* per-species DAG bootstraps / ATE folds / CV folds.

For example, 8 cores with 4 species → 4 species in parallel × 2
intra-species workers each.

## Resource log

Each run appends to `output_dir/resource_log.csv` with one row per step
per species: timestamp, species, step name, elapsed seconds, peak RAM
(MiB; via `peakRAM` if installed, else `gc()`-derived). Plot it with
the `perf-profile` vignette to find the slowest stages.

## Overrides at the call site

You can override any YAML setting from R without touching the file:

```{r}
cast_run_from_config(
  "my_run.yaml",
  cv_k = 10,                     # override cast_batch_args.cv_k
  models = c("cast", "rf")       # subset models
)
```

This is handy for quick sensitivity checks; for the canonical run, edit
the YAML so the change is captured in version control.
