---
title: "Profiling Multi-species Runs with `resource_log.csv`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profiling Multi-species Runs with resource_log.csv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE,
                      fig.width = 9, fig.height = 5)
```

## Where the time goes

Every `cast_batch()` (or `cast_run_from_config()`) run appends one row
per species x pipeline-step to `<output_dir>/resource_log.csv`. Steps
are: `prepare`, `dag`, `select`, `fit`, `eval`, `cv`, `predict`,
`cate`, `consistency`. Columns:

```
timestamp, species, step, elapsed_sec, peak_mb
```

The vignette below uses base R + ggplot2 to show the two views I find
most useful: a **stacked bar** of cumulative wall time per species, and
a **dot plot** of peak RAM per step. No hidden helper -- copy and adapt.

## Read & tidy

```{r}
log_path <- file.path("castSDM_output", "resource_log.csv")
log <- read.csv(log_path, stringsAsFactors = FALSE)
log$step <- factor(
  log$step,
  levels = c("prepare", "dag", "select", "fit",
             "eval", "cv", "predict", "cate", "consistency")
)
head(log)
```

If a step ran multiple times (e.g. you re-ran with `resume = TRUE`),
keep only the latest row per (species, step):

```{r}
log <- log[order(log$timestamp), ]
log <- log[!duplicated(log[, c("species", "step")], fromLast = TRUE), ]
```

## Per-species elapsed time

```{r}
library(ggplot2)
ggplot(log, aes(x = species, y = elapsed_sec, fill = step)) +
  geom_col() +
  scale_fill_brewer(palette = "Set3") +
  labs(x = NULL, y = "Wall-clock seconds",
       title = "castSDM pipeline cost by species and step") +
  coord_flip() +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")
```

The stacked bars make it obvious which species dominate the budget and
which steps inside each species do most of the work -- usually `dag`
(bootstrap structure learning) or `cv` (k-fold refits).

## Peak RAM per step

```{r}
ggplot(log, aes(x = step, y = peak_mb, color = species)) +
  geom_jitter(width = 0.18, size = 2.5, alpha = 0.85) +
  labs(x = NULL, y = "Peak RAM (MiB)",
       title = "Memory headroom by pipeline step") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

If `peakRAM` was not installed when the run executed, `peak_mb` is the
total `gc()`-reported maximum since each step's reset -- an upper bound
that includes prior allocations. Install `peakRAM` for tighter numbers:

```{r}
install.packages("peakRAM")
```

## Quick targets for tuning

* `dag` taking >> the rest -- lower `dag_R` (bootstrap replicates) or
  `dag_max_rows` (subsample cap).
* `cv` taking >> the rest -- run with `parallel = TRUE` once a worker
  budget is set.
* High `peak_mb` on `predict` -- switch to `cast_predict_tiled()` with a
  modest `tile_size`.
* Highly variable `cv` cost across species -- usually a presence-count
  effect; the rare ones are picking up `esm` paths or hitting RF
  internal parallelism limits.

## Drop-in helper

If you call this often, wrap the read/normalise/plot into a script
local to your project. Keep it short -- the columns are stable, and a
20-line script you understand is more useful than a packaged plot you
have to look up.
